How does the writing system get informative? Differentiation or conservation?

### 2023-08-21 manuscript/wp4.tex
ms 20/08

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}. To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission + Communication.

@@

We recruited 287 participants via the Prolific platform. Participants were paid £2.00 for participation plus additional bonuses of up to £1.08 as detailed below (median bonus: £0.74). The median completion time was 15~m with a median hourly rate of £8.05 (£10.94 including bonus). We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2). 14 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks. A further three participants were lost to communication-game pairing failures. The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{(A) Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. (B) Transmission + Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was culturally transmitted. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for the three unseen items. Transmission continued for nine generations in each of the ten independent chains.

@@

\subsubsection{Test procedure in Transmission + Communication}

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit on an independently selected set of seen items). The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission + Communication condition in Experiment~1 (differentiation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items. There are some isolated examples of differentiation through implicit generalization (K5, M1, P5, S2) and explicit innovation (K9, L7, M7, O4, Q7).}
	\label{dif_com}
	\end{figure}

The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}. Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O). In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6). Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable. In particular, there are two main kinds of color-expressive system that emerge. The first are the generalization-based expressive systems: K5, M1, P5, and S2. In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape. For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>. In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->). We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.

The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4. In M7, for example, the participant added <r>, <g>, and <b> (presumably red, gold, blue) to the ends of the words, although the participant's partner only reciprocated the <g> spelling and seemingly failed to understand what was meant by <r> and <b>. In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9. Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that express both shape and color simultaneously. Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color. In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-blue>, <-oran>, <-pink>, <-red>, <-yell>), but these pairs were excluded and replaced with a new pair before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words. However, we allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red). Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography. But this did not happen, even under the communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure. So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color. Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion} \textit{transmission}, and \textit{invention}).

@@

The methods were largely identical to Experiment~1 with two major exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.

@@

We recruited 297 native-English participants via the Prolific platform. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78). The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus). 17 participants were excluded because they (or their partners) used English color words or failed the auditory attention checks. A further 10 participants were lost to communication-game pairing failures. Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission + Communication condition (10 chains of 9 pairs of subjects).

@@

The alien objects and word stems were identical to Experiment~1. Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that uses three distinct suffixes to systematically express each of the colors. These suffixes were made~up~of a consonant from the set \{/f/, /s/, /ʃ/\} and a vowel from the set \{/ə/, /ɛɪ/, /əʊ/\}, but the exact forms were randomized for each chain (e.g., one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/). The initial orthographic system was transparent and generated based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}. The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ in speech or that /ʃ/ might be spelled <s> in writing). We attempted to achieve this balance by combining consonants that are very similar with vowels that are very dissimilar, while also avoiding reuse of any sounds present in the stems. The spoken forms were synthesized using the Apple text-to-speech synthesizer (Moira voice).

@@

As in Experiment~1, each transmission chain was run for nine generations. Additionally, the nine generations were divided into three epochs, each lasting three generations. During Epoch~I (Generations~1 to~3), the three spoken suffixes were distinct (as described above), allowing the spoken language to express all three colors without ambiguity. During Epoch~II (Generations~4 to~6), the spoken language had two distinct suffix forms, reducing its informativeness. During Epoch~III (Generations~7 to~9), all three spoken suffixes were homophonous, just as in Experiment~1, making the spoken language entirely uninformative about color. This was achieved through two sets of sound changes, the first occurring in the transition from Epoch~I to~II and the second occurring in the transition from Epoch~II to~III. An example is illustrated in Fig.~\ref{sound_change}. In the first sound change, two of the spoken suffix forms were chosen at random and the consonant from one (chosen at random) was paired with the vowel from the other, resulting in a new suffix form that replaced the original two. For example, a language that uses the suffixes /fəʊ/, /ʃə/, /sɛɪ/ in Epoch~I, might use the suffixes /fɛɪ/, /ʃə/, /fɛɪ/ in Epoch~II (i.e., /s/→/f/ and /əʊ/→/ɛɪ/). In the second sound change, the remaining two suffix forms were merged in the same way, resulting in full homophony. For example, in Epoch~III, the suffixes might become /fə/, /fə/, /fə/ (i.e., /ʃ/→/f/ and /ɛɪ/→/ə/). Crucially, the spellings do not automatically change following a sound change event; rather, the orthographic system is free to adapt (or not) in response to the sound changes.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_lrn.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission-only condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items.}
	\label{con_lrn}
	\end{figure}

The results for all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission + Communication condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items.}
	\label{con_com}
	\end{figure}

The results for the Transmission + Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}.

@@

>>>>

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}. To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.

@@

We recruited 287 participants via the Prolific platform. Participants were paid £2.00 for participation plus additional bonuses of up to £1.08 as detailed below (median bonus: £0.74). The median completion time was 15~m with a median hourly rate of £8.05 (£10.94 including bonus). We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2). 14 participants were excluded because they (or their partners) used English color words (8) or failed the auditory attention checks (6). A further three participants were lost to communication-game pairing failures. The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{(A) Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. (B) Transmission~+ Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was culturally transmitted. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items. Transmission continued for nine generations in each of ten independent chains. In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend). The role of the secondary participant is to act as a communicative partner for the primary participant, inducing pressure for the language to become informative.

@@

\subsubsection{Test procedure in Transmission~+ Communication}

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit on an independently selected set of seen items). The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/dif_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission~+ Communication condition in Experiment~1 (differentiation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items. There are some isolated examples of differentiation through implicit generalization (K5, M1, P5, S2) and explicit innovation (K9, L7, M7, O4, Q7).}
	\label{dif_com}
	\end{figure}

The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}. Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O). In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6). Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable. In particular, there are two main kinds of color-expressive system that emerge. The first are the generalization-based expressive systems: K5, M1, P5, and S2. In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape. For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>. In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->). We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.

The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4. In M7, for example, the participant added <r>, <g>, and <b> (presumably red, gold, blue) to the ends of the words, although the participant's partner only reciprocated the <g> spelling and seemingly failed to understand what was meant by <r> and <b>. In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9. Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that express both shape and color simultaneously. Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color. In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-pink> or <-red>, <-oran> or <-yell>, and <-blue>), but these pairs were excluded and replaced before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words. We allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red). Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography. But this did not happen, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure. So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color. Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).

@@

The methods were identical to Experiment~1 with two main exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.

@@

The experiment was completed by 297 native-English participants recruited through Prolific. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78). The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus). 17 participants were excluded because they (or their partners) failed the auditory attention checks (15) or used English color words (2). A further 10 participants were lost to communication-game pairing failures. Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).

@@

The alien objects and word stems were identical to Experiment~1. Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that uses three distinct suffixes to systematically express each of the colors. A separate set of suffixes was created for each chain by concatenating a randomly drawn consonant from \{/f/, /s/, /ʃ/\} and a randomly drawn vowel from \{/ə/, /ɛɪ/, /əʊ/\}, both without replacement. For example, one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/. The initial orthographic system was transparent and based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}. The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ or /ʃ/ in speech or that /ʃ/ might be spelled <s> or <x> in writing). We attempted to achieve this balance by combining consonants that are very similar with vowels that are very dissimilar, while also avoiding reuse of any sounds present in the stems. The spoken forms were synthesized using the Apple text-to-speech synthesizer (Moira voice).

@@

As in Experiment~1, each transmission chain was run for nine generations. Additionally, the nine generations were divided into three epochs, each lasting three generations. During Epoch~I (Generations~1 to~3), the three spoken suffixes were distinct (as described above), allowing the spoken language to express all three colors without ambiguity. During Epoch~II (Generations~4 to~6), the spoken language had two distinct suffix forms, reducing its informativeness. During Epoch~III (Generations~7 to~9), all three spoken suffixes were homophonous, just as in Experiment~1, making the spoken language entirely uninformative about color. This was achieved through two sets of sound changes, the first occurring in the transition from Epoch~I to~II and the second occurring in the transition from Epoch~II to~III. An example is illustrated in Fig.~\ref{sound_change}. In the first sound change, two of the spoken suffix forms were chosen at random and the consonant from one (chosen at random) was paired with the vowel from the other, resulting in a new suffix form that replaced the original two. In the second sound change, the remaining two suffix forms were merged in the same way, resulting in full homophony. Crucially, the spellings do not automatically change following a sound change event; rather, the orthographic system is free to adapt (or not) in response to the sound changes. Note also that an individual participant does not directly experience the sound changes; rather, a Generation~4 participant, for example, would always hear Epoch~II sounds, while observing spellings produced by a Generation~3 participant (presumably representing the Epoch~I sounds). In reality, of course, sound change is more gradual, with individual speakers experiencing both outgoing and incoming spoken forms within their lifetimes.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_lrn.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission-only condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled A--J and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items. The final systems are characterized by the conservation of form without the conservation of expressivity.}
	\label{con_lrn}
	\end{figure}

The results for all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{con_lrn}. It is immediately clear that color-expressive suffixes (as indicated by vertical stripes) are maintained fairly reliably through the first epoch; perfectly in the case of Chains~A, B, D, F, and~I, and with some errors in the other five chains. Some of these errors are very minor, such as the use of <sha> instead of <xa> for one item in H3, while others are more catastrophic, such as the early loss of the <sei>/<xa> distinction in Chain~J. Generation~4 represents the first real test of the orthographic systems in the face of sound change, and, in most cases, the Generation~3 systems are preserved quite faithfully (Chains~A, B, D, E, H, and~I), but by the end of the epoch (Generation~6), many have degenerated into redundant systems that encode shape (Chains~B, G, and~J) or transparent systems that mirror the Epoch~II pattern of homophony (Chains~C, E, F, and~I). These processes continue into Epoch~III and by the ninth generation, all systems have become degenerate, redundant, or some mixture of the two. The one exception is Chain~D, whose original spellings---<sei>, <fa>, <xoe>---were conserved perfectly through to the final generation with only one generalization error in Generation~8, which was quickly reverted in Generation~9.

Like Experiment~1, redundant systems are characteristic of the Transmission-only condition, especially in Epoch~III. As the spoken suffixes becomes more homophonous, the variant spellings are increasingly conditioned on shape rather than color, perhaps because the spoken language signals to learners that the language does not mark color, so they rationalize the system as three words with idiosyncratic spellings. Interestingly, however, all chains exhibited conservation of spelling \textit{form}, even if the way in which spelling was conditioned on meaning was lost. For example, Chain~A ultimately represents the sound /fə/ with the spellings <foe> and <xa>, spellings that are internally inconsistent and contrary to standard uses of the Roman alphabet, but which trace their origins back to the original seed orthography. Overall, then, the Transmission-only condition in Experiment~2 is characterized by the conservation of spelling \textit{form} without the conservation of expressivity.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission~+ Communication condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items.}
	\label{con_com}
	\end{figure}

The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{con_com}. Like Transmission-only, the seed systems are mostly maintained faithfully through Epoch~I; perfectly in the case of Chains~K, L, N, O, and~J, and with some errors in the other five chains, although some of these errors are non-destructive changes, such as <x> being replaced with <sh> in R1. Several systems were then maintained through Epoch~II, notably Chains~O, P, R, and~T, and, in one case, through to the end of Epoch~III (Chain~R, albeit with the original <xei> spelling replaced with <shei>). The Chain~O system was preserved up to Generation~7, Chain~P was maintained up to Generation~7 and almost to Generation~9 with two modifications (<sha> instead of <xa> and <fa> instead of <fei>), and Chain~T was conserved faithfully up to Generation~8. The final form of Chain~Q was also fully expressive, albeit through a combination of both conservation and differentiation: The <fei> form was conserved from the seed orthography, the <oxie> spelling appears to derive from a misremembering of <xoe> (partial conservation), and the <fe> spelling, which began as a typo introduced in Q6, seems to have been generalized across the blue items in Q7, perhaps to differentiate them from the yellow items (indeed, the participant's partner made the same generalization). A similar case of differentiation might also have occurred in L9, where the <sol> spelling (originally a typo on <so>) was generalized across the yellow items, resulting in a semi-expressive system (although the participant's partner generalized the <sol> spelling across shape).

Unlike Experiment~1, no participant pairs attempted to use English color letters and there was only one case of a pair using English color words in Generation~9 of Chain~K, although this pair was excluded and replaced (this generation was the only case in which the training input was fully degenerate, which would have resulted in a strong pressure to find a communicative solution in the form of English). Presumably, the general conservation of expressive spelling in Experiment~2 negated the need to innovate novel systems.

@@

In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.
